% !TEX root = ../thesis.tex


\acknowledgements{%
  First and foremost, I would like to thank Prof.~Hadzibabic for the opportunity to work in his research group during this year. Although I spent less than half of my time in Cambridge working in the Cavendish due to the Covid-19 pandemic, it was a great learning experience. I want to thank my colleagues, primarily Nishant Dogra for proofreading this thesis at the last minute and for giving me a lot of helpful tips along the way as well as Julian Schmitt and Christoph Eigen who have helped me many times and were always there to answer my questions. I am even more grateful for the independence I was given while I was working in Cambridge and I hope that the results of this thesis can help in building up our new ultracold atom experiment over the coming years. 
  
  Thank you to AMD for making the Ryzen 7 3700X processor, its 8 cores with 16 threads and \SI{36}{MB} of cache provided me with the processing power necessary to perform many of the simulations done for this thesis. Countless nights of silent and efficient computation were spent and without this CPU, this thesis would not have been possible\footnote{I am not sponsored by AMD.}.

  I want to thank my friends from the Cambridge University Dancesport Team and my dance partner Erika for the amazing time we had while we could still dance. Being a part of the best university dance team in the UK has been a unique experience and the stress of working towards a competition was always a welcome distraction from the stress of working in the laboratory.
  Thank you to Leon, who, although having studied at the other place for a long time before coming to Cambridge, is a great guy and an amazing friend.
  Thank you to my friends, Jorgos and Clara for the fun evenings (and sometimes nights) we spent playing Minecraft together. 

  For all your continued support over the years, for your love and constant encouragement, I want to thank my family. Thank you to my mother, for the most warm welcomes when I visited Bonn, my father, for always believing in me, may he read this thesis from his seat in heaven. Thank you to my brother, for being the most awesome brother I ever had.

  From the bottom of my heart, I want to thank my girlfriend Johanna. I cannot recall how many hours we spent talking to each other every week. You always had two open ears for me, you listened to my problems and helped me overcome my doubts. I love you and I look forward to being with you again.
}
